An Extension of LTL with Rules\\ and its Application to Runtime Verification

### 2019-02-02 rule-ltl.tex
fixing problems

\section{Extending Propositional LTL}

\subsubsection{Quantified LTL}

@@

\subsubsection{Rule-based LTL}

@@

We define models for FLTL based on {\em temporal} relations~\cite{Chomicki}. That is,
their last parameter
is a natural number. For a relation $R$, $R [ i ]$ is the relation obtained from $R$ by
restricting it to the value $i$ in the last parameter. For simplicity, we will describe henceforth
the logic with relations $R$ that have exactly two parameters
, hence $R [ i ]$ will have just one parameter whose domain will be denoted as $d ( R )$; this domain is fixed over all the value of the second parameter $i$. The the definition of the logic over other number of parameters is quite straightforward, and our implementation of runtime verification fully supports it.

@@

The formulas of the core FLTL logic are 
defined by the following grammar,
where $p$ denotes to a relation $R$,
$a$ is a constant representing a value in $d ( R )$ and $x$ is a variable with $\domain ( x ) = d ( p )$.
\begin{center}
$\varphi ::= \true  \; | \;
    p ( a ) \; | \;
    p ( x ) \; | \;
    ( \varphi \wedge \varphi ) \;  |   \;
   \neg \varphi \; | \;
   \bigcirc \varphi \; | \; 
   ( \varphi \; \until \; \varphi ) \; | $ \\
   $ \ominus \; \varphi \; | \;
    ( \varphi  \; \since  \; \varphi ) \; | \;
    \exists x \; \varphi$
\end{center}

@@

Additional operators are defined as in the propositional logic. In addition we define
$\forall x \; \varphi = \neg \exists x \neg \varphi$.
Restricting the modal operators to the past ones
($\since$, $\ominus$ and the ones obtained from them) 
forms the logic PFLTL.
The quantification over values of variables here is `static' in the sense that
the assigned value is independent of the state in the execution. We
denote static quantification with $\exists$ and $\forall$, while
dynamic quantification uses $\Exists$ and $\Forall$.

@@

A model is a set of
 relations ${\cal R} = \{ R_1 \, \ldots , R_m \}$, where each of these relations has a last parameter $i$ that is a natural number. 
% 
It it standard, and more intuitive (in particular for runtime verification) to cast models of temporal logic in terms of sequences states rather than temporal relations. To do that,
 we denote by $R_k [ i ]$ the restriction of
 the relation $R_k$ to the value $i$ of the last parameter. A model ${\cal R}$ is then a sequence ${\cal R} [ 1 ]\, {\cal R} [ 2 ] \ldots$, where
 each {\em state} ${\cal R} [ i ] = {\cal R} [i] = \{ R_1 [i], \ldots , R_m [ i ] \}$  (a set of relations).
 The denotation $p_k$ will correspond to the relation $R_k$.

@@

\noindent {\bf Example 1.}
Let $p$ and $q$ be temporal reations, where their
restrictions to the state $s_i$ is $p[i]$ and $q[i]$, respectively, are unary relations $q(x)$ and $p(x)$. The
specification that we want to monitor is that for each value
$a$, $p(a)$ appears in all the states where $q(a)$ has appeared an even number of times so far (for the odd occurrences, $p(a)$ can also appear, but does not have to). This is an extension of
Wolper's example over LTL, carried over to the first-order case. To show that this is not expressible in FLTL (and PFLTL),
consider models (executions) where only one data element $a$ appears. We can prove that in this case, by a simple structural induction on (sub)formulas, that for each property $\varphi$
we can replace occurrences of variables within
predicates by $a$, i.e., $p(x)$ and $q(x)$ become $p(a)$
and $q  (a )$, respectively, and that we can throw away
the quantification ($\forall x$, $\exists x$), obtaining
an equivalent formula. This is then equivalent to
an LTL formula that is obtained by replacing
the occurrences of $p(a)$ and $q(a)$ by propositions
$p_a$ and $p_b$, respectively. However, by~\cite{Wolper},
such a formula cannot express the said property.

This example demonstrates that the so called ``non-counting''~\cite{Thomas} property of LTL carries over
to FLTL. This lack of expressiveness can be resolved by
adding propositional variables (i.e., predicates that are
only state dependent) with dynamic quantification over them.
Using parametric automata as a specification formalism, as in~\cite{Grum,havelund-rv-data-2018,Meredith2011,Reger2015} can also express this property.

\noindent {\bf Example 2.}
%%%%

Extending FLTL (PFLTL) with quantifiers over relations, we obtain QFLTL
(and the past-restricted version QPFLTL, respectively). 
The syntax includes $\Exists p \, \varphi$, 
where $p$ denotes an auxiliary relation.
We also allow $\Forall p \varphi = \neg \Exists \neg \varphi$. The semantics is as follows.

@@

>>>>

\section{Extending Propositional LTL}
\label{sec:extending-prop-ltl}

\subsubsection{Extending with Dynamic Quantification}

@@

\subsubsection{Extending with Rules}

@@

We define models for FLTL based on {\em temporal} relations~\cite{Chomicki}. That is,
their last parameter
is a natural number. For a relation $R$, $R [ i ]$ is the relation obtained from $R$ by
restricting it to the value $i$ in the last parameter. For simplicity, we will describe henceforth
the logic with relations $R$ that have exactly two parameters, 
hence $R [ i ]$ will have just one parameter whose domain will be denoted as $d ( R )$; this domain is fixed over all the values of the second parameter $i$. The definition of the logic over other numbers of parameters is quite straightforward, and our implementation of runtime verification fully supports it.

\klaus{Above, what about instead of writing: ``this domain is fixed over all the values of the second parameter $i$'' to write:
``this domain is independent of the values of the second parameter $i$'' - if that is what you mean to say.}

\klaus{Above: we need to say explicitly what $i$ intuitively is meant to represent: positions, time, ....}

@@

The formulas of the core FLTL logic are 
defined by the following grammar,
where $p$ denotes a relation $R$,
$a$ is a constant representing a value in $d ( R )$ and $x$ is a variable with $\domain ( x ) = d ( p )$.
\begin{center}
$\varphi ::= \true  \; | \;
    p ( a ) \; | \;
    p ( x ) \; | \;
    ( \varphi \wedge \varphi ) \;  |   \;
   \neg \varphi \; | \;
   \bigcirc \varphi \; | \; 
   ( \varphi \; \until \; \varphi ) \; | $ %\\
   $ \ominus \; \varphi \; | \;
    ( \varphi  \; \since  \; \varphi ) \; | \;
    \exists x \; \varphi$
\end{center}

@@

Additional operators are defined as in the propositional logic. In addition we define
$\forall x \; \varphi = \neg \exists x \neg \varphi$.
Restricting the modal operators to the past ones
($\since$, $\ominus$ and the ones obtained from them) 
forms the logic PFLTL.
The quantification over values of variables here is `static' in the sense that values quantified over are independent of the state in the execution. We
denote static quantification with $\exists$ and $\forall$, while
dynamic quantification is expressed with $\Exists$ and $\Forall$
(see e.g. Section \ref{sec:extending-prop-ltl}).

@@

A model is a set of
 relations ${\cal R} = \{ R_1 \, \ldots , R_m \}$, where each of these relations has a last parameter $i$ that is a natural number. 
% 
It is standard, and more intuitive (in particular for runtime verification) to cast models of temporal logic in terms of sequences states rather than temporal relations. To do that,
 we denote by $R_k [ i ]$ the restriction of
 the relation $R_k$ to the value $i$ of the last parameter. A model ${\cal R}$ is then a sequence ${\cal R} [ 1 ]\, {\cal R} [ 2 ] \ldots$, where
 each {\em state} ${\cal R} [i] = \{ R_1 [i], \ldots , R_m [ i ] \}$  (a set of relations).
 The denotation $p_k$ will correspond to the relation $R_k$.

\klaus{Above, did you not already say the following before:
``we denote by $R_k [ i ]$ the restriction of
 the relation $R_k$ to the value $i$ of the last parameter.''.}

@@

\noindent {\bf Example.}
Let $p$ and $q$ be temporal relations, where their
restrictions to the state $s_i$, $p[i]$ and $q[i]$, respectively, are unary relations $p(x)$ and $q(x)$. The
specification that we want to monitor is that for each value
$a$, $p(a)$ appears in all the states where $q(a)$ has appeared an even number of times so far (for the odd occurrences, $p(a)$ can also appear, but does not have to). This is an extension of
Wolper's example over LTL, carried over to the first-order case. To show that this is not expressible in FLTL (and PFLTL),
consider models (executions) where only one data element $a$ appears. We can prove that in this case, by a simple structural induction on (sub)formulas, that for each property $\varphi$
we can replace occurrences of variables within
predicates by $a$, i.e., $p(x)$ and $q(x)$ become $p(a)$
and $q  (a )$, respectively, and that we can throw away
the quantification ($\forall x$, $\exists x$), obtaining
an equivalent formula. This is then equivalent to
an LTL formula that is obtained by replacing
the occurrences of $p(a)$ and $q(a)$ by propositions
$p_a$ and $p_b$, respectively. However, by~\cite{Wolper},
such a formula cannot express the said property.
This example demonstrates that the so called ``non-counting''~\cite{Thomas} property of LTL carries over
to FLTL.

\subsubsection{Extending with Quantification}

This lack of expressiveness can be resolved by
adding propositional variables (i.e., predicates that are
only state dependent) with dynamic quantification over them.
Using parametric automata as a specification formalism, as in~\cite{Grum,havelund-rv-data-2018,Meredith2011,Reger2015} can also express this property.

Extending FLTL (PFLTL) with quantifiers over relations, we obtain QFLTL
(and the past-restricted version QPFLTL, respectively). 
The syntax includes $\Exists p \, \varphi$, 
where $p$ denotes an auxiliary relation.
We also allow $\Forall p\ \varphi = \neg \Exists \neg \varphi$. The semantics is as follows.

@@

\subsubsection{Extending with Rules}
